A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-08-24 model.tex
edited summary
minor changes

If the $\p_r\rhog$ can be neglected in Eq. \ref{spm}, then instability
requires 
\begin{align}
  \left|\frac{\p\ln{\epsilon}}{\p r}\right| \gtrsim
  \frac{1}{\Htilde}\frac{\left(1+\epsilon\right)}{\sqrt{\epsilon}}. 
\end{align}
That is, if the dust-to-gas ratio varies on a radial lengthscale less
than $O(\Htilde)$, i.e. its vertical lengthscale,  
%$O(\widetilde{H}\sqrt{\epsilon})$ for $\epsilon\ll 1$, or less than
%$O(\widetilde{H}/\sqrt{\epsilon})$ for $\epsilon\gg 1$, 
then the system is potentially unstable.

@@

\subsection{Thermodynamic view of dusty instabilities}\label{dust_work}
Now consider constant $c_s$ but $\tstop\neq0$. 
%When $c_s(r,z)$ is non-uniform but $\tstop=0$, 
Eq. \ref{int_rel} indicates that $\sigma^2$ is generally complex, so
we may have growing oscillations. Writing $\sigma = -\omega + \ii s$
where $\omega$ and $s$ are the real frequency and growth rates, 
respectively, the imaginary part of Eq. \ref{int_rel} gives 
\begin{align}
  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\int\rho\left(|\dd
    v_r|^2 + |\dd v_z|^2\right)dV}, \label{thermal_instability}
\end{align}
%In this case 
%the effective energy equation is 
%\begin{align}\label{dusty_cooling}
%  \frac{DP}{Dt} &= - P \nabla \cdot \bm{v} + \mathcal{C}
%\end{align} 
%The function $\mathcal{C}$ can be considered as a cooling term. 
%By
%analogy with enabling VSI in an adiabatic gas through cooling
%\citep{nelson13,lin15}, we may expect a `dusty VSI' to develop 
%if this term is large enough (physically meaning diffusion of the
%dust-fraction is strong). 
%Note that $\mathcal{C}$ is also responsible for the streaming
%instability  in the strong-drag limit \citep{laibe14}, though this
%does not require vertical stratification. In the form of
%Eq. \ref{dusty_cooling}, however, the streaming instability has a
%thermodynamical interpretation, similar to thermal instabilities in
%stars \citep{latter06}.  
which represent instabilities due to dust-gas friction. 
%Our formalism
%gives these dusty instabilities a thermodynamical interpretation. 
% We further discuss this concept in \S\ref{dust_as_thermo}. 
%In \S\ref{dust_work} we showed that instabilities may 
%arise from the term proportional $\imag\left( \nabla\cdot\dd\bm{v}^*
%\delta C\right)$. 
 Since drag ($\delta \mathcal{C}$) appears as a source term in our
effective energy equation, the quantity 
$\imag\left[\left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}\right]$
may be thought of as a correlation between compression/expansion and  
heating/cooling.

>>>>

We can neglect density gradients in Eq. \ref{spm} 
if $r\p_r\ln{\rhog}\sim O(1)$ and $H_\epsilon/\Hgas\gg
\sqrt{\epsilon}\hgas$. For example, if $\epsilon\simeq 0.01$ and
$\hgas\simeq 0.05$, then we require $H_\epsilon/\Hg\gg
5\times10^{-3}$. (So the dust layer thickness $\Hd\gg 5
\times10^{-3}\Hgas$ as well.) Then 
instability requires 
\begin{align}
  \left|\frac{\p\ln{\epsilon}}{\p r}\right| \gtrsim
  \frac{1}{\Htilde}\frac{\left(1+\epsilon\right)}{\sqrt{\epsilon}}. 
\end{align}
That is, if the dust-to-gas ratio varies on a radial lengthscale less
than $O(\Htilde)$, i.e. its vertical lengthscale,  
%$O(\widetilde{H}\sqrt{\epsilon})$ for $\epsilon\ll 1$, or less than
%$O(\widetilde{H}/\sqrt{\epsilon})$ for $\epsilon\gg 1$, 
then the system is potentially unstable.

@@

\subsection{Thermodynamics of dusty instabilities}\label{dust_work}
Now consider constant $c_s$ but $\tstop\neq0$. 
%When $c_s(r,z)$ is non-uniform but $\tstop=0$, 
Eq. \ref{int_rel} indicates that $\sigma^2$ is generally complex, so
we may have growing oscillations. Writing $\sigma = -\omega + \ii s$
where $\omega$ and $s$ are the real frequency and growth rates, 
respectively, the imaginary part of Eq. \ref{int_rel} gives 
\begin{align}
  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\int\rho\left(|\dd
    v_r|^2 + |\dd v_z|^2\right)dV}, \label{thermal_instability}
\end{align}
%In this case 
%the effective energy equation is 
%\begin{align}\label{dusty_cooling}
%  \frac{DP}{Dt} &= - P \nabla \cdot \bm{v} + \mathcal{C}
%\end{align} 
%The function $\mathcal{C}$ can be considered as a cooling term. 
%By
%analogy with enabling VSI in an adiabatic gas through cooling
%\citep{nelson13,lin15}, we may expect a `dusty VSI' to develop 
%if this term is large enough (physically meaning diffusion of the
%dust-fraction is strong). 
%Note that $\mathcal{C}$ is also responsible for the streaming
%instability  in the strong-drag limit \citep{laibe14}, though this
%does not require vertical stratification. In the form of
%Eq. \ref{dusty_cooling}, however, the streaming instability has a
%thermodynamical interpretation, similar to thermal instabilities in
%stars \citep{latter06}.  
which represent instabilities due to dust-gas friction. 
%Our formalism
%gives these dusty instabilities a thermodynamical interpretation. 
% We further discuss this concept in \S\ref{dust_as_thermo}. 
%In \S\ref{dust_work} we showed that instabilities may 
%arise from the term proportional $\imag\left( \nabla\cdot\dd\bm{v}^*
%\delta C\right)$. 
 Since drag ($\delta \mathcal{C}$) appears as a source term in our
effective energy equation, the quantity 
$\imag\left[\left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}\right]$
may be thought of as a correlation between compression/expansion and  
heating/cooling.

### 2016-08-24 summary.tex
minor changes

Fig. \ref{compare_eigenvals_ts0d02} shows the unstable modes found for
this setup. We find the usual VSI modes, here with $\omega> 3,\, s<
0.2$. However, we also find the mode marked by the blue asterisk with
frequency 
\begin{align*}
\sigma = (0.3631\ii - 0.1251)\hgas\OmK. 
\end{align*}
This growth rate corresponds to $8.8$ orbits, comparable to the $12$
orbits reported by \citeauthor{loren15}. We visualize this mode in
Fig. \ref{result2d_loren}. We find this mode 
persists when $q=0$, so it is not the VSI. 
Using Eq. \ref{streaming_dispersion} with the above parameters,  
we also estimate a much smaller growth 
rate for the streaming instability of 
$O(10^{-3}\hgas\OmK)$, We thus also rule out
SI, as did \citeauthor{loren15}. However, the mode disappears with
$\tstop=0$, so it is associated with finite dust-gas drag.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_eigenvals_ts0d02}
  \caption{Unstable modes in a locally isothermal, dusty disk
    with stopping time  
    $\tstop=0.02\OmK^{-1}$. Modes with $\omega>3$ are due to
    VSI. The mode marked by the blue asterisk, visualized in
    Fig. \ref{result2d_loren}, is associated with finite dust-gas
    drag.   
    \label{compare_eigenvals_ts0d02}
  }
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/result2d_loren.ps}
  \caption{The mode marked by the blue asterisk in
    Fig. \ref{compare_eigenvals_ts0d02}. This is a possible eigenmode
    of the dusty instability discovered by  
    \cite{loren15} in their numerical simulations. The instability growth timescale and 
    radial spacing of the `toroidal vortices' are similar to that 
    reported by \citeauthor{loren15}. This instability is driven by
    dust-gas drag. The color scale shows the 
    perturbed  
    dust-to-gas ratio, $\delta\epsilon$; and the arrows show
    $\sqrt{\rho}\left(\dd v_x, \dd v_z\right)$.
    \label{result2d_loren}
  }
\end{figure}

This preliminary calculation only serves to highlight the possibility
of new axisymmetric dust-drag instabilities in stratified disks. One 
issue is that the base state for the above calculation is not a true
equilibrium (although \citeauthor {loren15} find the dust settling
time is much longer than the instability growth time). We will study
this instability in more detail with a proper basic
state in a follow up work.

>>>>

Fig. \ref{compare_eigenvals_ts0d02} shows the unstable modes found for
this setup. We find the usual VSI modes, here with $\omega> 3,\, s<
0.2$. However, we also find modes with $\omega<s$. One of these,
marked by the blue asterisk, has 
frequency 
\begin{align*}
\sigma = (0.3631\ii - 0.1251)\hgas\OmK. 
\end{align*}
This growth rate corresponds to $8.8$ orbits, comparable to the $12$
orbits reported by \citeauthor{loren15}. We visualize this mode in
Fig. \ref{result2d_loren}. We find this mode 
persists when $q=0$, so it is not the VSI. 
Using Eq. \ref{streaming_dispersion} with the above parameters,  
we also estimate a much smaller growth 
rate for the streaming instability of 
$O(10^{-3}\hgas\OmK)$, We thus also rule out
SI, as did \citeauthor{loren15}. However, the mode disappears with
$\tstop=0$, so it is associated with finite dust-gas drag.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_eigenvals_ts0d02}
  \caption{Unstable modes in a locally isothermal, dusty disk
    with stopping time  
    $\tstop=0.02\OmK^{-1}$. Modes with $\omega>3$ are due to
    VSI. The mode marked by the blue asterisk, visualized in
    Fig. \protect\ref{result2d_loren}
    , is associated with finite dust-gas
    drag. 
  }\label{compare_eigenvals_ts0d02}
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/result2d_loren.ps}
  \caption{Visualization of the mode marked by the blue asterisk in
    Fig. \protect\ref{compare_eigenvals_ts0d02}. 
    This is a possible eigenmode of the dusty instability discovered by  
    \cite{loren15} in their numerical simulations. The instability growth timescale and 
    radial spacing of the `toroidal vortices' are similar to that 
    reported by \citeauthor{loren15}. This instability is driven by
    dust-gas drag. The color scale shows the 
    perturbed  
    dust-to-gas ratio, $\delta\epsilon$; and the arrows show
    $\sqrt{\rho}\left(\dd v_x, \dd v_z\right)$. 
  }\label{result2d_loren}
\end{figure}

This preliminary calculation only serves to highlight the possibility
of new axisymmetric dust-drag instabilities in stratified disks. One 
issue is that the base state for the above calculation is not a true
equilibrium (although \citeauthor {loren15} find the dust settling
time is longer than the instability growth time). We will study
this instability in more detail with a proper basic
state in a follow up work.
